R for Clinical Study Reports and Submission

### 2021-09-19 tlf-disposition.Rmd
- avoid 'we' in writing a text or book - Prefix R verbs (good programming practices) so the user sees which packages are used quickly without reading text + ensure correct package is used eg filter/lag - There are standard ADaM variables that capture discontinuation reason EOSSTT DCSREAS DCSREASP - participants in the study, not patients - represent treatments by actual names - calculation of count+percent is improved (show count and percent in same column, when count is 0 no percent should be shown)

# Table for Study Patients

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize accounting of all patients who entered the study, 
in the section 10.1, disposition of patients of a CSR.

@@

In this chapter, we illustrate how to create a disposition table as below.

@@

The first step is to read relevant datasets into R. 
For disposition table, all the required information is saved in the ADSL dataset.
We can use `haven` package to read the dataset.

```{r}
adsl <- read_sas("adam_data/adsl.sas7bdat")
```

We illustrate how to prepare a report data for a simplified disposition 
of patients table using variables below:

- USUBJID: Unique subject identifier 
- TRT01P: Planed treatment
- TRT01PN: Planned treatment encoding
- DISCONFL: Discontinued from study flag
- DCREASCD: Discontinued from study reason

```{r}
adsl %>%
  select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
  head(4)
```
In the code below, we calculated patients in the analysis population.

```{r}
n_rand <- adsl %>%
  group_by(TRT01PN) %>%
  summarize(n = n()) %>%
  pivot_wider(
    names_from = TRT01PN,
    names_prefix = "n_",
    values_from = n
  ) %>%
  mutate(row = "Patients in population")
n_rand
```
In the code below, we calculate number and percentage of patients who discontinued the study.

Here we use `formatC()` to customize the numeric value to be 1 decimal with fixed width of 5.

```{r}
n_disc <- adsl %>%
  group_by(TRT01PN) %>%
  summarize(
    n = sum(DISCONFL == "Y"),
    pct = formatC(n / n() * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  pivot_wider(
    names_from = TRT01PN,
    values_from = c(n, pct)
  ) %>%
  mutate(row = "Discontinued")

n_disc
```
In the code below, we calculate number and percentage of patients 
who completed/discontinued the study in different reasons.

```{r}
n_reason <- adsl %>%
  group_by(TRT01PN) %>%
  mutate(n_total = n()) %>%
  group_by(TRT01PN, DCREASCD) %>%
  summarize(
    n = n(),
    pct = formatC(n / unique(n_total) * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  pivot_wider(
    id_cols = DCREASCD,
    names_from = TRT01PN,
    values_from = c(n, pct),
    values_fill = list(n = 0, pct = "  0.0")
  ) %>%
  rename(row = DCREASCD)

@@

In the code below, we calculate number and percentage of patients 
who complete the study. We split `n_reason` because we need to customize 
the row order of the table.

```{r}
n_complete <- n_reason %>% filter(row == "Completed")

n_complete
```
In the code below, we calculate number and percentage of patients 
who discontinued the study in different reasons. 
Here we use `paste0("    ", row)` to add some leading space for individual reasons for study discontinuation.

```{r}
n_reason <- n_reason %>%
  filter(row != "Completed") %>%
  mutate(row = paste0("    ", row))

@@

Now we combined individual rows into the whole table for reporting purpose. 
`tbl_disp` is used as input for `r2rtf` to create final report.

```{r}
tbl_disp <- bind_rows(n_rand, n_complete, n_disc, n_reason) %>%
  select(row, ends_with(c("_0", "_54", "_81")))

@@

We start to define the format of the output. We highlighted items that 
is not discussed in previous discussion.

`rtf_title` define table title. We can provide a vector for the title argument. 
Each value is a separate line. 
The format can also be controlled by providing a vector input in text format.

@@

  # Table title
  rtf_title("Disposition of Patients") %>%

@@

>>>>

# Disposition {#disposition}

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), a summary needs to be provided that accounts for all participants who entered the study within Section 10.1, Disposition of participants.
This section summarizes the number of participants that were randomized, entered and completed each phase of the study. In addition, the reasons for all post-randomization discontinuations, grouped by treatment and by major reason of discontinuation are required (lost to follow-up, adverse event, ...).

@@

In this chapter, the creation of a typical disposition table is shown.

@@

The first step is to read in the relevant datasets into R.
For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
The dataset is provided in SAS7bdat format, which is the current standard in current clinical trial reporting and analysis.The `haven` package is able to read in the dataset, while maintaining its attributes (eg. variable labels).

```{r}
adsl <- haven::read_sas("adam_data/adsl.sas7bdat")
```

Following variables are used in the preparation of the disposition report:

-   USUBJID: Unique subject identifier
-   TRT01P: Planed treatment
-   TRT01PN: Planned treatment numeric encoding
-   DISCONFL: Discontinued from study flag
-   DCREASCD: Discontinued from Study reason coded

```{r}
adsl %>%
  dplyr::select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
  utils::head(4)
```

In the code below, the number of participants randomized is calculated.

```{r}
n_rand <- adsl %>%
  dplyr::group_by(TRT01P) %>%
  dplyr::summarize(n = n()) %>%
  tidyr::pivot_wider(
    names_from = TRT01P,
    names_prefix = "",
    values_from = n
  ) %>%
  dplyr::mutate(row = "participants Randomized")

n_rand
```

In the code below, the number of participants that discontinued the study is calculated.

Here we use the `formatC()` function to customize the numeric value to be 1 decimal with fixed width of 5.

```{r}
n_disc <- adsl %>%
  dplyr::group_by(TRT01P) %>%
  dplyr::summarize(
    n = base::sum(DISCONFL == "Y"),
    pct = base::formatC(n / n() * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  dplyr:: mutate(npct = base::paste0(n, " (", pct, "%)")) %>%
  dplyr::select(-c(n,pct)) %>%
  tidyr::pivot_wider(
    names_from = TRT01P,
    values_from = npct,
    values_fill = list(npct = "0")
  ) %>%
  dplyr::mutate(row = "Discontinued")

n_disc
```

To calculate the number and percentage of participants that completed/discontinued the study for each discontinuation reason, below code can be used.

```{r}
n_reason <- adsl %>%
  dplyr::group_by(TRT01P) %>%
  dplyr::mutate(n_total = dplyr::n()) %>%
  dplyr::group_by(TRT01P, DCREASCD) %>%
  dplyr::summarize(
    n = dplyr::n(),
    pct = base::formatC(n / unique(n_total) * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  dplyr:: mutate(npct = base::paste0(n, " (", pct, "%)")) %>%
  tidyr::pivot_wider(
    id_cols = DCREASCD,
    names_from = TRT01P,
    values_from = npct,
    values_fill = list(npct = "0")
  ) %>%
  dplyr::rename(row = DCREASCD)

@@

From the step above, the number and percentage of participants that completed the study can be isolated so change the
roworder in the final table.

```{r}
n_complete <- n_reason %>% 
  dplyr::filter(row == "Completed")

n_complete
```

For display purposes leading spaces are added to study discontinuation. This will mimick identation in the final report.

```{r}
n_reason <- n_reason %>%
  dplyr::filter(row != "Completed") %>%
  dplyr::mutate(row = paste0("    ", row))

@@

In the final step, the individual rows are combined and used as input for `r2rtf.

```{r}
tbl_disp <- base::rbind(n_rand, n_complete, n_disc, n_reason)

@@

In the below code, formatting of the final table is defined. Sections are highlighted that were not discussed before.

@@

  # Table title
  rtf_title("Disposition of participants") %>%

@@

The procedure to generate a disposition table can be summarized as follow:

-   Step 1: Read subject level data into R, i.e. `adsl`.
-   Step 2: Count participants in the analysis population and name the dataset `n_rand`.
-   Step 3: Count the number and percentage of participants that discontinued the study, and name the dataset `n_disc`.
-   Step 4: Count the number and percentage of participants that discontinued the study for different reasons, and name the dataset `n_reason`.
-   Step 5: Count the number and percentage of participants that completed the study, and name the dataset `n_complete`.
-   Step 6: Bind `n_rand`, `n_disc`, `n_reason` and `n_complete` by row. 
-   Step 7: Write the final table to RTF
